Improving the prospects for multimessenger astronomy with \earlywarning\ detection of compact binary coalescence.

### 2011-05-05 introduction.tex
First pass on introduction
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@743 ccdd05d5-56c1-461a-bf98-85f6876523b4

%
The coalescence of compact binary systems consisting of neutron stars (NS)
and/or black holes (BH) is the most promising source of gravitational radiation
for ground-based gravitational-wave detectors such as Advanced
\LIGO~\cite{ALIGOWeb}, Virgo~\cite{AVirgoWeb}, \GEO~\cite{GEOWeb}, and
\LCGT~\cite{LCGTWeb}.  Detections will constrain the mass and spin of compact
objects~\cite{finn1993,Poisson:1995ef}, the neutron star equation of
state~\cite{flanagan:021502,Read:2009}, general relativity in the strong-field
regime~\cite{Will:2005va}, the graviton
mass~\cite{PhysRevD.80.044002,Keppel:2010qu}, and the Lorentz Invariance
principle~\cite{Ellis2006402}.

Radio observation of binary pulsar systems in the Galaxy provide indirect
evidence that such systems are losing energy according to the prediction of
gravitational-wave emission from general
relativity~\cite{Taylor:1982,Abadie:2010p10836}.  One can estimate the expected
coalescence rate by extrapolating the orbital periods of systems observed
electromagnetically.  Although this procedure has large uncertainty, tens of
binary coalescence events are realistically expected to be observed in the
advanced detector era later this decade~\cite{Abadie:2010p10836}.

As a binary system loses energy to gravitational waves, its orbital separation
decreases. This causes a run-away inspiral with the gravitational-wave
amplitude and frequency increasing until the system eventually merges near the
innermost stable circular orbit (\ISCO).
%
\footnote{$f_\mathrm{ISCO} \approx 4.4\times 10^3\,$Hz$\,\times(M/\Msun)^{-1}$. 
for a test particle orbiting a Schwarzschild black hole with the total mass of
the system $M$.}
%
If a neutron star is involved it may become tidally disrupted near the merger.
This can provide the matter required to produce bright electromagnetic
counterparts in the moments following the merger~\cite{shibata:2007}.
\editorial{nvf: Chad thinks we need more mechanisms. I think this stands on its
own without the acceleration mechanism.} In order to observe the prompt and
potentially brightest electromagnetic emission, telescopes will have to be
pointing in the direction of the event during the seconds surrounding the
merger. In order to assure that this occurs more often than by random chance
from various astronomical surveys operating independently, the
gravitational-wave community has initiated a project to send alerts when
potential gravitational-wave transients are observed.

In 2010, \textsc{ligo} completed its sixth science run (S6) and Virgo
completed its third science run (VSR3). When both \textsc{ligo} and Virgo were
operating, several all-sky detection pipelines operated in a low-latency
configuration~\cite{HugheyGWPAW2011, S6lowlatency}. Candidates were sent for
electromagnetic followup to several telescopes including Swift,
\textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008, HugheyGWPAW2011}.
The S6 analyses achieved latencies of $~\sim30$--$60$ minutes, which were
dominated by a human vetting process. The intrinsic gravitational-wave trigger
generator latency was $\sim$2--5 minutes.  In this work we will assume that
the human vetting process can be automated in the advanced detector era and
that the dominating latency will be trigger generation. From this perspective
we will try to address the realistic expectations for reducing trigger latency
from $\sim$ minutes \emph{after} the merger to minutes \emph{before} the
merger for sufficiently loud events.  Figure \ref{fig:earlywarning} presents
the possibility for early detection of NS--NS binaries with Advanced \LIGO\
assuming the rate predictions in \cite{Abadie:2010p10836}.
%
\begin{figure}
\includegraphics{N_before_Tc.pdf}
\caption{\label{fig:earlywarning} Expected number of NS--NS sources that will
be detectable $T_{bc}$ seconds before coalescence.  The solid line is the most
likely yearly rate estimate $N_{\mathrm{re}}$ for Advanced \LIGO\ and the
shaded region is the interval $N_{\mathrm{low}}$ to $N_{\mathrm{high}}$ from
\cite{Abadie:2010p10836}.  Note that assuming \SNR\ 8 is sufficient for
detection and that we observe $N_{\mathrm{re}} = 40$ events per year with a
detector having the ZERO\_DET\_high\_P noise model described
in~\cite{ALIGONoiseZERO_DET_high_P}, $\sim10$ sources may be detected within 10
seconds of merger and $\sim1$ sources maybe detected within 100 seconds of
merger.  Other noise models, ZERO\_DET\_low\_P~\cite{ALIGONoiseZERO_DET_low_P}
and BHBH\_20deg~\cite{ALIGONoiseBHBH_20deg} provide better results for early
detection but at the cost of fewer total events observed above SNR 8.}
\end{figure}

>>>>

%
\begin{comment}
The coalescence of compact binary systems consisting of neutron stars (NS)
and/or black holes (BH) is the most promising source of gravitational radiation
for Advanced \LIGO~\cite{ALIGOWeb}, Virgo~\cite{AVirgoWeb}, \GEO~\cite{GEOWeb}, and
\LCGT~\cite{LCGTWeb}.  Tens of
binary coalescence events are expected to be observed in the
advanced detector era later this decade~\cite{Abadie:2010p10836}.
\end{comment}

As a binary system loses energy to gravitational waves, its orbital separation
decreases. This causes a run-away inspiral with the gravitational-wave
amplitude and frequency increasing until the system eventually merges near the
innermost stable circular orbit (\ISCO).  If a neutron star is involved it may become tidally disrupted near the merger.
This disrupted matter can fuel a bright electromagnetic
counterpart in the system's final moments as a binary~\cite{shibata:2007}.
In order to observe the prompt and
potentially most intense electromagnetic emission, telescopes must point during the seconds surrounding the
merger.

\editorial{Citation needed for LOOC-UP} To make this possible, the gravitational-wave community initiated a project to send alerts when
potential gravitational-wave transients are observed.  In October 2010, \textsc{ligo} completed its sixth science run (S6) and Virgo
completed its third science run (VSR3).  While both \textsc{ligo} detectors and Virgo were
operating, several all-sky detection pipelines operated in a low-latency
configuration, namely \textsc{mbta}, ihope, Coherent WaveBurst, and Omega~\cite{HugheyGWPAW2011, S6lowlatency}. \editorial{Get references for these low-latency pipelines.} The S6 analysis achieved latencies of 30--60 minutes, which were
dominated by a human vetting process. Candidates were sent for
electromagnetic followup to several telescopes; Swift,
\textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008, HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved the best gravitational-wave trigger
generation latencies of 2--5 minutes.  We assume that
in the advanced detector era the vetting process will be automated, so current
trigger generation and telescope actuation would then dominate latency.

To this end, we have the ambition of reporting candidates not minutes \emph{after} the merger, but seconds \emph{before}.  By looking for threshold crossings before the signal leaves the detection band, it is possible to trade some signal to noise ratio (\textsc{snr}) for latency.  Figure \ref{fig:earlywarning} shows projected early trigger rates for NS--NS binaries in Advanced \LIGO\
assuming the event rate predictions in \cite{Abadie:2010p10836}.
%
\begin{figure}
\includegraphics{N_before_Tc.pdf}
\caption{\label{fig:earlywarning} Expected number of NS--NS sources that will
be detectable $T_{bc}$ seconds before coalescence.  The solid line is the most
likely yearly rate estimate $N_{\mathrm{re}}$ for Advanced \LIGO\ and the
shaded region is the interval $N_{\mathrm{low}}$ to $N_{\mathrm{high}}$ from
\cite{Abadie:2010p10836}.  Note that assuming \SNR\ 8 is sufficient for
detection and that we observe $N_{\mathrm{re}} = 40$ events per year with a
detector having the ZERO\_DET\_high\_P noise model described
in~\cite{ALIGONoiseZERO_DET_high_P}, $\sim10$ sources may be detected within 10
seconds of merger and $\sim1$ sources maybe detected within 100 seconds of
merger.  Other noise models, ZERO\_DET\_low\_P~\cite{ALIGONoiseZERO_DET_low_P}
and BHBH\_20deg~\cite{ALIGONoiseBHBH_20deg} provide better results for early
detection but at the cost of fewer total events observed above SNR 8.}
\end{figure}
